---
title: "R Notebook"
output: html_notebook
---

```{r setup, include = FALSE}
library(tidyverse)
library(afex)

# return a vector with a specified correlation to a pre-existing vector
rnormpre <- function (x, rho=0, ymean=0, ysd=1) {
  n <- length(x)
  y <- stats::rnorm(n)
  z <- rho * scale(x)[,1] + sqrt(1 - rho^2) * 
    scale(stats::resid(stats::lm(y ~ x)))[,1]
  yresult <- ymean + ysd * z
  yresult
}
```

## Set study parameters

The means and variance in the intercepts and slopes can be estimated from pilot data.

```{r}
# number of subjects and items
sub_n <- 100
item_n <- 50

grand_int <- 2.5 # overall mean (intercept)
grand_sd <- 1.2 # overall SD

sub_sd <- 0.1 # SD of subject intercepts

item_dist <- c(.1,.2,.4,.2,.1) # distribution of item ratings

cor_mean <- .2 # mean correlation between ratings and traits
cor_sd <- .1 # sd of subjects' correlations between ratings and traits
```

## Simulate data

Make random intercepts (trait rating) for each item.

```{r}

item <- tibble(
  item_id = paste0("I",1:item_n),
  item_i = sample(1:length(item_dist), item_n, replace = T, prob = item_dist)
)

ggplot(item) + 
  geom_histogram(
    aes(item_i), 
    binwidth = 1, 
    fill="white", 
    color="red"
  )
```

Make random intercepts (mean rating) and slopes (correlation between traits and ratings) for each subject.

```{r}

sub <- tibble(
  sub_id = paste0("S", 1:sub_n),
  sub_i = rnorm(sub_n, 0, sub_sd),
  sub_r = rnorm(sub_n, cor_mean, cor_sd) %>% pmax(-1, .) %>% pmin(1, .)
)

ggplot(sub) + 
  geom_histogram(
    aes(sub_r), 
    binwidth = 0.05, 
    fill="white", 
    color="blue"
  )

ggplot(sub) + 
  geom_histogram(
    aes(sub_i), 
    binwidth = 0.2, 
    fill="white", 
    color="red"
  )
```


## Simulate Data

```{r}
for (i in 1:sub_n) {
  r <- sub[[i,3]]
  sub_i <- sub[[i,2]]
  id <- sub[[i,1]]
  ratings <- rnormpre(item$item_i, r, grand_int + sub_i, grand_sd) %>% 
    round() %>%
    pmax(1, .) %>% # set the minimum value to 1
    pmin(5, .) # set the maximum value to 5
  
  item[id] = ratings
}
```


## tidy data
```{r}
dat <- item %>%
  gather(sub_id, rating, 3:ncol(.)) %>%
  rename(trait = item_i)
```


## Visualise by-subject and overall trait-rating correlations
```{r, fig.width = 8, fig.height = 6}
dat %>%
  group_by(sub_id) %>%
  mutate(r = cor(trait, rating)) %>%
  ungroup() %>%
  ggplot() +
  geom_smooth(aes(trait, rating, color = r, group = sub_id), 
              method = "lm", se = FALSE) +
  geom_smooth(aes(trait, rating), method = "lm", color = "black", size = 3) +
  scale_color_viridis_c(alpha = 0.5)
```



## LMER
```{r}
m <- lmer(rating ~ trait + 
            (1 | item_id) + 
            (1 + trait | sub_id), 
          data = dat)

summary(m)
```


Confirm that random slopes for trait by subject are actually correlated with the sub_r the data were modelled from.

```{r}
rand_slope <- ranef(m)$sub_id %>%
  as.tibble(rownames = "sub_id") %>%
  left_join(sub, by = "sub_id")

cor(rand_slope$sub_r, rand_slope$trait)
```



